Halving the cost of quantum addition

### 2017-09-26 paper.tex
First pass of modifications to meet nature quantum information - Separated latex used to generate figures into separate files - Changed section names - Added required sections

\begin{abstract}
We improve the number of T gates needed to perform an $n$-bit adder from $8n + O(1)$ \citep{Amy2013, Cuccaro2004, AustinDiscussionsAndEmails2017} to $4n + O(1)$.
We do so via a ``temporary logical-AND" construction, which uses four T gates to store the logical-AND of two qubits into an ancilla and zero T gates to later erase the ancilla.

Temporary logical-ANDs are a generally useful tool when optimizing T-counts.
They can be applied to integer arithmetic, modular arithmetic, rotation synthesis, the quantum Fourier transform, Shor's algorithm, Grover oracles, and many other circuits.
Because T gates dominate the cost of quantum computation based on the surface code, and temporary logical-ANDs are widely applicable, our constructions represent a significant reduction in projected costs of quantum computation.

We also present an $n$-bit controlled adder circuit with T-count of $8n + O(1)$, a temporary adder that can be computed for the same cost as the normal adder but whose result can be kept until it is later uncomputed without using T gates, and discuss some other constructions whose T-count is improved by the temporary logical-AND.
\end{abstract}

@@

\section{Introduction}
\label{sec:introduction}

@@

\begin{figure}
  \resizebox{\linewidth}{!}{
    \Qcircuit @R=0.7em @C=0.7em {
      &i_0 &&\ctrl{1} &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\ctrl{1}&\ctrl{1}&\qw &i_0     &&&&&\\
      &t_0 &&\ctrl{1} &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\ctrl{1}&\targ   &\qw &&&(t+i)_0 &&&\\
      &    &&         &\ctrl{2}&\qw     &\ctrl{3}&\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\ctrl{3}&\qw     &\ctrl{1}&\qw     &        &    &&&        &&&\\
      &i_1 &&\qw      &\targ   &\ctrl{1}&\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\ctrl{1}&\targ   &\qw     &\ctrl{1}&\qw &i_1     &&&&&\\
      &t_1 &&\qw      &\targ   &\ctrl{1}&\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\ctrl{1}&\qw     &\qw     &\targ   &\qw &&&(t+i)_1 &&&\\
      &    &&         &        &        &\targ   &\ctrl{2}&\qw     &\ctrl{3}&\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\ctrl{3}&\qw     &\ctrl{1}&\targ   &\qw     &        &        &        &    &&&        &&&\\
      &i_2 &&\qw      &\qw     &\qw     &\qw     &\targ   &\ctrl{1}&\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\ctrl{1}&\targ   &\qw     &\qw     &\qw     &\qw     &\ctrl{1}&\qw &i_2     &&&&&\\
      &t_2 &&\qw      &\qw     &\qw     &\qw     &\targ   &\ctrl{1}&\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\ctrl{1}&\qw     &\qw     &\qw     &\qw     &\qw     &\targ   &\qw &&&(t+i)_2 &&&\\
      &    &&         &        &        &        &        &        &\targ   &\ctrl{2}&\qw     &\ctrl{3}&\qw     &\ctrl{3}&\qw     &\ctrl{1}&\targ   &\qw     &        &        &        &        &        &        &    &&&        &&&\\
      &i_3 &&\qw      &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\targ   &\ctrl{1}&\qw     &\qw     &\qw     &\ctrl{1}&\targ   &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\ctrl{1}&\qw &i_3     &&&&&\\
      &t_3 &&\qw      &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\targ   &\ctrl{1}&\qw     &\qw     &\qw     &\ctrl{1}&\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\targ   &\qw &&&(t+i)_3 &&&\\
      &    &&         &        &        &        &        &        &        &        &        &\targ   &\ctrl{2}&\targ   &\qw     &        &        &        &        &        &        &        &        &        &    &&&        &&&\\
      &i_4 &&\qw      &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\ctrl{1}&\qw &i_4     &&&&&\\
      &t_4 &&\qw      &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\targ   &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\qw     &\targ   &\qw &&&(t+i)_4 &&&\\
    }
  }
  \caption{
	A 5-bit adder with T-count of 16.
	Uses Clifford operations, four logical-AND computations each with a T-count of 4, and four logical-AND erasures requiring no T gates.
	Generalizes to an $n$-bit adder with a T-count of $4n - 4$.
	See \autoref{fig:full-adder-block} for the adder building-block and \autoref{fig:temporary-logical-AND} for the logical-AND computation and uncomputation circuits.
  }
  \label{fig:multi-bit-adder-example}
\end{figure}

Our paper is divided into four sections.
We discuss existing work and motivate the need to optimize T-counts in \autoref{sec:introduction}.
\autoref{sec:construction} presents our $4n + O(1)$ T-count adder and explains its pieces, including the temporary logical-AND.
In \autoref{sec:other-applications} we justify the assertion that our constructions represent widely applicable optimizations by giving examples of quantum tasks and cost projections that are improved by efficient adders and temporary logical-ANDs.
Finally, \autoref{sec:conclusion} summarizes our contributions and discusses future work.

\section{Adder Construction}
\label{sec:construction}

@@

\begin{figure}
  \resizebox{\linewidth}{!}{
    \Qcircuit @R=0.7em @C=0.7em {
      &c_k &&\ctrl{2}&\qw     &\ctrl{3}&\qw &\qw &\qw    &\qw &\qw &\qw &\qw    &\qw &\qw &\qw &\qw    &\qw &\qw &\ctrl{3}&\qw     &\ctrl{1}&\qw     &\qw &c_k     &&&&& \\
      &i_k &&\targ   &\ctrl{1}&\qw     &\qw &\qw &\qw    &\qw &\qw &\qw &\qw    &\qw &\qw &\qw &\qw    &\qw &\qw &\qw     &\ctrl{1}&\targ   &\ctrl{1}&\qw &i_k     &&&&& \\
      &t_k &&\targ   &\ctrl{1}&\qw     &\qw &\qw &\qw    &\qw &\qw &\qw &\qw    &\qw &\qw &\qw &\qw    &\qw &\qw &\qw     &\ctrl{1}&\qw     &\targ   &\qw &&&(t+i)_k &&& \\
      &    &&        &        &\targ   &\qw &    &c_{k+1}&    &    &    &\ldots &    &    &    &c_{k+1}&    &    &\targ   &\qw     &        &        &    &&&        &&& \\
    }
  }
  \caption{
	Our adder circuit building-block, with a T-count of 4.
	A variant of the Cuccaro adder building-block \citep{Cuccaro2004}.
  }
  \label{fig:full-adder-block}
\end{figure}

@@

\section{Other Applications}
\label{sec:other-applications}

@@

\begin{figure}
  \resizebox{\linewidth}{!}{
    \Qcircuit @R=0.7em @C=0.7em {
      &x &&\ctrl{1} &\qw &x  && &&          &&\ctrl{2}&\qw     &\targ    &\qTi &\targ     &\qw &\qw &\qw &\\
      &y &&\ctrl{1} &\qw &y  &&=&&          &&\qw     &\ctrl{1}&\targ    &\qTi &\targ     &\qw &\qw &\qw &\\
      &  &&         &\qw &xy && &&|A\rangle &&\targ   &\targ   &\ctrl{-2}&\qT  &\ctrl{-2} &\qH &\qS &\qw &\\
    }
  }
  \vspace*{0.25cm} \\
  \resizebox{\linewidth}{!}{
    \Qcircuit @R=0.7em @C=0.7em {
      &x  &&\ctrl{1} &\qw &x && &&&\qw &\qw    &\ctrl{1}         &\qw &&&&&&&&&&\\
      &y  &&\ctrl{1} &\qw &y &&=&&&\qw &\qw    &\gate{Z}         &\qw &&&&&&&&&&\\
      &xy &&\qw      &    &  && &&&\qH &\meter &\cw \cwx \bullet &    &&&&&&&&&&\\
    }
  }
  \caption{
	How to compute and uncompute the logical-AND of two qubits.
	The computation circuit (top) has a T-count of 4 and a T-depth of 1.
	Note that the $|A\rangle$ state input contributes to the T-count, because $|A\rangle$ states are the resource used to perform T gates.
	The uncomputation circuit (bottom) uses a measure-and-fixup approach \cite{Jones2013} that requires only Clifford gates, and so has a T-count of zero.
    \\
    An alternative uncomputation construction is to simply do the reverse of the computation circuit.
    This alternative approach has a net T-count of 2 (because an $|A\rangle$ state is recovered).
    The resulting temporary logical-AND with a T-count of 6 would still be an improvement on existing work, but would be inferior to the measure-and-fixup approach shown above.
  }
  \label{fig:temporary-logical-AND}
\end{figure}

@@

\begin{figure}
  (a) \resizebox{\linewidth}{!}{
    \Qcircuit @R=0.7em @C=0.7em {
      &&\text{control} &&&&\qw     &\qw     &\qw     &\qw &\qw &\qw    &\qw &\qw &\qw &\qw    &\qw &\qw &\qw &\qw    &\qw &\qw &\qw     &\qw     &\ctrl{2}&\qw     &\ctrl{2}&\qw     &\qw &&&\text{control}               &&&&&&\\
      &&&c_k            &&&\ctrl{3}&\qw     &\ctrl{4}&\qw &\qw &\qw    &\qw &\qw &\qw &\qw    &\qw &\qw &\qw &\qw    &\qw &\qw &\ctrl{4}&\qw     &\qw     &\qw     &\qw     &\ctrl{3}&\qw &c_k                          &&&&&&&&\\
      &&&i_k            &&&\targ   &\ctrl{2}&\qw     &\qw &\qw &\qw    &\qw &\qw &\qw &\qw    &\qw &\qw &\qw &\qw    &\qw &\qw &\qw     &\ctrl{2}&\ctrl{1}&\qw     &\ctrl{1}&\targ   &\qw &i_k                          &&&&&&&&\\
      &&               &&&&        &        &        &    &    &       &    &    &    &       &    &    &    &       &    &    &        &        &        &\ctrl{1}&\qw     &        &    &    &&&&&&&&\\
      &&&t_k            &&&\targ   &\ctrl{1}&\qw     &\qw &\qw &\qw    &\qw &\qw &\qw &\qw    &\qw &\qw &\qw &\qw    &\qw &\qw &\qw     &\ctrl{1}&\qw     &\targ   &\qw     &\targ   &\qw &&&&&(t+i \cdot \text{control})_k &&&&&\\
      &&               &&&&        &        &\targ   &\qw &    &c_{k+1}&    &    &    &\ldots &    &    &    &c_{k+1}&    &    &\targ   &\qw     &        &        &        &        &    &    &&&&&&&&\\
    }
  }
  \vspace*{0.25cm} \\
  (b) \resizebox{\linewidth}{!}{
    \Qcircuit @R=0.7em @C=0.7em {
      &a &&\ctrl{2}&\qw     &\qw     &\ctrl{3}&\qw &a          &&&&&&&&&&&&&&&&&&&&&&\\
      &b &&\targ   &\ctrl{1}&\ctrl{1}&\targ   &\qw &b          &&&&&&&&&&&&&&&&&&&&&&\\
      &c &&\targ   &\ctrl{1}&\targ   &\qw     &\qw &&&&(a+b+c)_0  &&&&&&&&&&&&&&&&&&&\\
      &  &&        &        &\qw     &\targ   &\qw &&&&(a+b+c)_1  &&&&&&&&&&&&&&&&&&&\\
    }
  }
  \caption{
	Variations on our adder construction.
	(a) is a controlled-adder building-block with a T-count of 8.
	The addition will only occur when the control is on.
	(b) is a temporary-adder building-block with a T-count of 4.
	Has the same T-count as a normal addition, but makes the result available before any uncomputation steps are needed.
  }
  \label{fig:variations-on-full-adder-block}
\end{figure}

@@

\begin{figure}
  \resizebox{\linewidth}{!}{
    \Qcircuit @R=0.7em @C=0.7em {
      &\ustick{n}&\diagup\qw &\multigate{3}{U}&\qw     &\qw &&               &&&&\ustick{n}&\diagup\qw &\multigate{3}{U}&\qw     &\qw &          &&            &&\ustick{n}&\diagup\qw &\qw      &\multigate{3}{U}&\qw      &\qw     &\qw &\\
      &          &\ctrl{1}   &\ghost{U}       &\ctrl{1}&\qw &&               &&&&          &\ctrl{1}   &\ghost{U}       &\ctrl{1}&\qw &          &&            &&          &\ctrl{1}   &\qw      &\ghost{U}       &\qw      &\ctrl{1}&\qw &\\
      &          &\ctrl{1}   &\ghost{U}       &\ctrl{1}&\qw &&\stackrel{?}{=}&&&&          &\ctrl{2}   &\ghost{U}       &\ctrl{2}&\qw &          &&\Rightarrow &&          &\ctrl{2}   &\qw      &\ghost{U}       &\qw      &\ctrl{2}&\qw &\\
      &          &\targ      &\ghost{U}       &\targ   &\qw &&               &&&&          &\targ      &\ghost{U}       &\targ   &\qw &          &&            &&          &\qw        &\targ    &\ghost{U}       &\targ    &\qw     &\qw &\\
      &          &           &                &        &    &&               &&&&\qO       &\targ      &\qw             &\targ   &\qw &|0\rangle &&            &&          &           &\ctrl{-1}&\qw             &\ctrl{-1}&\qw     &    &\\
    }
  }
  \caption{
	A sufficient condition for replacing a pair of Toffoli gates with a temporary logical-AND, saving 4 T gates.\\
	1) The later Toffoli gate must be uncomputing the earlier Toffoli gate.\\
	2) Intermediate operations must not be sensitive to the presence of the entangled ancilla.
  }
  \label{fig:pair-definition}
\end{figure}

@@

\section{Conclusion}
\label{sec:conclusion}

@@

\begin{figure}
  \resizebox{\linewidth}{!}{
    \Qcircuit @R=0.7em @C=0.7em {
      &\qT &\qw && &&&\ctrl{2}&\qw     &\ctrl{3}&\qw     &\qw &\qw     &\ctrl{3}&\qw     &\ctrl{2}&\qw &\\
      &\qT &\qw &&=&&&\targ   &\ctrl{1}&\qw     &\ctrl{1}&\qw &\ctrl{1}&\qw     &\ctrl{1}&\targ   &\qw &\\
      &\qT &\qw && &&&\targ   &\ctrl{1}&\targ   &\targ   &\qT &\targ   &\targ   &\ctrl{1}&\targ   &\qw &\\
      &    &    && &&&        &        &\targ   &\qw     &\qS &\qw     &\targ   &\qw     &        &    &\\
    }
  }
  \caption{
	Three T gates can be synthesized by performing a temporary logical-AND, one T gate, and various Clifford operations.
	Clifford operations can't produce T gates, so the extra T gates must be coming from the temporary logical-AND.
	This proves that performing the temporary logical-AND requires at least two T gates.
  }
  \label{fig:lower-bound-logical-AND}
\end{figure}

@@

\section{Acknowledgements}

@@

>>>>

\begin{abstract}
We improve the number of T gates needed to perform an $n$-bit adder from $8n + O(1)$ \citep{Amy2013, Cuccaro2004, AustinDiscussionsAndEmails2017} to $4n + O(1)$.
We do so via a ``temporary logical-AND" construction, which uses four T gates to store the logical-AND of two qubits into an ancilla and zero T gates to later erase the ancilla.
Temporary logical-ANDs are a generally useful tool when optimizing T-counts.
They can be applied to integer arithmetic, modular arithmetic, rotation synthesis, the quantum Fourier transform, Shor's algorithm, Grover oracles, and many other circuits.
Because T gates dominate the cost of quantum computation based on the surface code, and temporary logical-ANDs are widely applicable, our constructions represent a significant reduction in projected costs of quantum computation.
We also present an $n$-bit controlled adder circuit with T-count of $8n + O(1)$, a temporary adder that can be computed for the same cost as the normal adder but whose result can be kept until it is later uncomputed without using T gates, and discuss some other constructions whose T-count is improved by the temporary logical-AND.
\end{abstract}

@@

\section*{Introduction}

@@

\begin{figure}
  \input{figures/adder.qcircuit}
  \caption{
	A 5-bit adder with T-count of 16.
	Uses Clifford operations, four logical-AND computations each with a T-count of 4, and four logical-AND erasures requiring no T gates.
	Generalizes to an $n$-bit adder with a T-count of $4n - 4$.
	See \autoref{fig:full-adder-block} for the adder building-block and \autoref{fig:temporary-logical-AND} for the logical-AND computation and uncomputation circuits.
  }
  \label{fig:multi-bit-adder-example}
\end{figure}

\section*{Results}

@@

\begin{figure}
  \input{figures/adder-building-block.qcircuit}
  \caption{
	Our adder circuit building-block, with a T-count of 4.
	A variant of the Cuccaro adder building-block \citep{Cuccaro2004}.
  }
  \label{fig:full-adder-block}
\end{figure}

@@

@@

\begin{figure}
  \input{figures/temporary-logical-and.qcircuit}
  \caption{
	How to compute and uncompute the logical-AND of two qubits.
	The computation circuit (top) has a T-count of 4 and a T-depth of 1.
	Note that the $|A\rangle$ state input contributes to the T-count, because $|A\rangle$ states are the resource used to perform T gates.
	The uncomputation circuit (bottom) uses a measure-and-fixup approach \cite{Jones2013} that requires only Clifford gates, and so has a T-count of zero.
    \\
    An alternative uncomputation construction is to simply do the reverse of the computation circuit.
    This alternative approach has a net T-count of 2 (because an $|A\rangle$ state is recovered).
    The resulting temporary logical-AND with a T-count of 6 would still be an improvement on existing work, but would be inferior to the measure-and-fixup approach shown above.
  }
  \label{fig:temporary-logical-AND}
\end{figure}

@@

\begin{figure}
  \input{figures/other-adder-building-blocks.qcircuit}
  \caption{
	Variations on our adder construction.
	(a) is a controlled-adder building-block with a T-count of 8.
	The addition will only occur when the control is on.
	(b) is a temporary-adder building-block with a T-count of 4.
	Has the same T-count as a normal addition, but makes the result available before any uncomputation steps are needed.
  }
  \label{fig:variations-on-full-adder-block}
\end{figure}

@@

\begin{figure}
  \input{figures/paired-toffoli-replacement-rule.qcircuit}
  \caption{
	A sufficient condition for replacing a pair of Toffoli gates with a temporary logical-AND, saving 4 T gates.\\
	1) The later Toffoli gate must be uncomputing the earlier Toffoli gate.\\
	2) Intermediate operations must not be sensitive to the presence of the entangled ancilla.
  }
  \label{fig:pair-definition}
\end{figure}

@@

\section*{Discussion}

@@

\begin{figure}
  \input{figures/synthesize-t-gates-from-temporary-logical-and.qcircuit}
  \caption{
	Three T gates can be synthesized by performing a temporary logical-AND, one T gate, and various Clifford operations.
	Clifford operations can't produce T gates, so the extra T gates must be coming from the temporary logical-AND.
	This proves that performing the temporary logical-AND requires at least two T gates.
  }
  \label{fig:lower-bound-logical-AND}
\end{figure}

@@

\section*{Methods}

The described circuits can be verified with a quantum circuit simulator, such as ProjectQ or LIQUID or Quirk.

\section*{Acknowledgements}

@@

\section*{Competing Interests}

No competing interests.

\section*{Contributions}

Craig Gidney discovered and refined the temporary logical-AND construction, applied it to adders and the other discussed circuits, and is the sole author of the paper.

\section*{Funding}

Work was done while employed at Google.
